Adaptive group-regularized logistic elastic net regression

### 2019-02-20 docs/manuscript.Rnw
added barplots of multiplier estimates to metabolomics_alzheimer.R example

@@

	\cite{li_bayesian_2010} show that (\ref{eq:elasticnetprior}) may be written as a computationally more convenient scale mixture of normals, with mixing parameter $\btau = \begin{bmatrix} \tau_1 & \cdots & \tau_p \end{bmatrix} \tr$. Using this result, we write the generalised elastic net model in its Bayesian form as:
	\begin{subequations}\label{eq:enmodel2}
		\begin{align}
		\y| \bbeta & \sim \prod_{i=1}^n \mathcal{B} \left( m_i, \expit ( \x\tr_i \bbeta ) \right),  \\
		\bbeta | \btau & \sim \prod_{g=1}^G \prod_{j \in \G(g)} \mathcal{N} \left(0,\frac{1}{\lambda' _g \lambda_2} \frac{\tau_j - 1}{\tau_j} \right), \\
		\btau & \sim \prod_{j=1}^p \mathcal{TG} \left( \frac{1}{2},\frac{8 \lambda_2}{\lambda_1^2}, \left(1,\infty \right) \right).
		\end{align}
	\end{subequations}
	Here, $\mathcal{TG} \( k,\theta,\( x_l,x_u\) \)$ denotes the truncated gamma distribution with shape $k$, scale $\theta$, and domain $\(x_l,x_u\)$. In this Bayesian formulation the penalty parameters $\blambda = \begin{bmatrix} \lambda_1 & \lambda_2 & \lambda'_1 & \cdots & \lambda'_G \end{bmatrix} \tr$ play the role of the hyperparameters in a Bayesian hierarchical model.

@@

	Many forms of empirical Bayes exist, the most formal one being maximisation of the marginal likelihood with respect to the hyperparameters. The resulting hyperparameter estimates are then plugged into the prior. The marginal likelihood is often introduced as a measure of model evidence given the observed data and is computed by integrating the product of likelihood and prior with respect to the model parameters. In the case of the elastic net introduced in (\ref{eq:enmodel2}) this results in the following marginal empirical Bayes posterior for $\bbeta$:
	\begin{align}
	p_{\hat{\blambda}}(\bbeta | \y) & = \frac{\L (\y ; \bbeta) \pi_{\hat{\blambda}} (\bbeta) }{p_{\hat{\blambda}}(\y)} = \frac{\int_{\btau} \L (\y ; \bbeta) \pi_{\hat{\blambda}}(\bbeta | \btau) \pi_{\hat{\blambda}}(\btau) \, d\btau}{p_{\hat{\blambda}}(\y)}, \nonumber \\
	\hat{\blambda} & := \underset{\blambda}{\argmax} \log p_{\blambda} (\y) = \underset{\blambda}{\argmax} \int_{\bbeta} \int_{\btau} \L_{\blambda}(\y, \bbeta, \btau) \, d\bbeta d\btau \nonumber \\
	& = \underset{\blambda}{\argmax} \int_{\bbeta} \int_{\btau} \L (\y ; \bbeta) \pi_{\blambda}(\bbeta | \btau) \pi_{\blambda}(\btau) \, d\bbeta d\btau. \label{eq:mmlestimate}
	\end{align}
	The integrals in (\ref{eq:mmlestimate}) are intractable in the case of the elastic net. In the omics setting the integrals are also high-dimensional, in which case numerical and Monte Carlo approximation methods become tedious and computationally expensive. Moreover, Laplace approximation is known to suffer from low accuracy in many high-dimensional settings \cite[]{shun_laplace_1995}. In \cite{casella_empirical_2001} an EM algorithm is described that estimates the hyperparameters. This EM algorithm iteratively maximises the expected joint log likelihood, such that the sequence:
	\begin{equation}\label{eq:ebEM}
	\blambda^{(k + 1)} = \underset{\blambda}{\argmax} \E_{\bbeta, \btau | \y} \[ \log \L_{\blambda}(\y, \bbeta, \btau) | \blambda^{(k)} \]
	\end{equation}
	converges to a local maximum of the marginal likelihood. The difficulty herein is in the calculation of the expected joint log likelihood. \cite{casella_empirical_2001} suggests to approximate the expectation by its Monte Carlo expectation. Although elegant and simple, this method requires a converged MCMC sample from the posterior for every iteration: a computationally intensive procedure. \cite{roy_selection_2017} introduce generalised importance sampling for the Bayesian elastic net, such that we just need a limited, pre-specified number of MCMC chains. However, this still requires several converged chains, so is not feasible in many high-dimensional omics settings. We propose to tackle this problem by approximating the expectation in (\ref{eq:ebEM}) using variational Bayes.

@@

	\section{Application to metabolomics in alzheimer}\label{sec:app}

<<lines_metabolomics_alzheimer_res1_auc, fig.cap="Cross-validated AUC in alzheimer example using metabolomics class co-data, with a) the full range and b) cut-out", out.width="100%", fig.asp=2/3>>=
@

@@

@@

<<lines_metabolomics_alzheimer_res4_auc, fig.cap="Cross-validated AUC in alzheimer example using all co-data, with a) the full range and b) cut-out", out.width="100%", fig.asp=2/3>>=
@

>>>>

% set options
<<settings, include=FALSE, echo=FALSE>>=
opts_knit$set(root.dir="..", base.dir="../figs")
opts_chunk$set(fig.align='center', fig.path="../figs/", 
               echo=FALSE, cache=FALSE, message=FALSE)
@

% read figure code
<<figures, include=FALSE>>=
read_chunk("code/figures.R")
@

@@

	\cite{li_bayesian_2010} show that (\ref{eq:elasticnetprior}) may be written as
	a computationally more convenient scale mixture of normals, with mixing 
	parameter $\btau = \begin{bmatrix} \tau_1 & \cdots & \tau_p \end{bmatrix} 
	\tr$. Using this result, we write the generalised elastic net model in its 
	Bayesian form as:
	\begin{subequations}\label{eq:enmodel2}
		\begin{align}
		\y| \bbeta & \sim \prod_{i=1}^n \mathcal{B} \left( m_i, \expit ( \x\tr_i 
		\bbeta ) \right),  \\
		\bbeta | \btau & \sim \prod_{g=1}^G \prod_{j \in \G(g)} \mathcal{N} \left(0
		,\frac{1}{\lambda' _g \lambda_2} \frac{\tau_j - 1}{\tau_j} \right), \\
		\btau & \sim \prod_{j=1}^p \mathcal{TG} \left( \frac{1}{2},\frac{8 \lambda_2
		}{\lambda_1^2}, \left(1,\infty \right) \right).
		\end{align}
	\end{subequations}
	Here, $\mathcal{TG} \( k,\theta,\( x_l,x_u\) \)$ denotes the truncated gamma 
	distribution with shape $k$, scale $\theta$, and domain $\(x_l,x_u\)$. In this
	Bayesian formulation the penalty parameters $\blambda = \begin{bmatrix} 
	\lambda_1 & \lambda_2 & \lambda'_1 & \cdots & \lambda'_G \end{bmatrix} \tr$ 
	play the role of the hyperparameters in a Bayesian hierarchical model.

@@

	Many forms of empirical Bayes exist, the most formal one being maximisation of
	the marginal likelihood with respect to the hyperparameters. The resulting 
	hyperparameter estimates are then plugged into the prior. The marginal 
	likelihood is often introduced as a measure of model evidence given the 
	observed data and is computed by integrating the product of likelihood and 
	prior with respect to the model parameters. In the case of the elastic net 
	introduced in (\ref{eq:enmodel2}) this results in the following marginal 
	empirical Bayes posterior for $\bbeta$:
	\begin{align}
	p_{\hat{\blambda}}(\bbeta | \y) & = \frac{\L (\y ; \bbeta) \pi_{\hat{\blambda}
	} (\bbeta) }{p_{\hat{\blambda}}(\y)} = \frac{\int_{\btau} \L (\y ; \bbeta) 
	\pi_{\hat{\blambda}}(\bbeta | \btau) \pi_{\hat{\blambda}}(\btau) \, d\btau}{p_
	{\hat{\blambda}}(\y)}, \nonumber \\
	\hat{\blambda} & := \underset{\blambda}{\argmax} \log p_{\blambda} (\y) = 
	\underset{\blambda}{\argmax} \int_{\bbeta} \int_{\btau} \L_{\blambda}(\y, 
	\bbeta, \btau) \, d\bbeta d\btau \nonumber \\
	& = \underset{\blambda}{\argmax} \int_{\bbeta} \int_{\btau} \L (\y ; \bbeta) 
	\pi_{\blambda}(\bbeta | \btau) \pi_{\blambda}(\btau) \, d\bbeta d\btau. 
	\label{eq:mmlestimate}
	\end{align}
	The integrals in (\ref{eq:mmlestimate}) are intractable in the case of the 
	elastic net. In the omics setting the integrals are also high-dimensional, in 
	which case numerical and Monte Carlo approximation methods become tedious and 
	computationally expensive. Moreover, Laplace approximation is known to suffer 
	from low accuracy in many high-dimensional settings 
	\cite[]{shun_laplace_1995}. In \cite{casella_empirical_2001} an EM algorithm 
	is described that estimates the hyperparameters. This EM algorithm iteratively
	maximises the expected joint log likelihood, such that the sequence:
	\begin{equation}\label{eq:ebEM}
	\blambda^{(k + 1)} = \underset{\blambda}{\argmax} \E_{\bbeta, \btau | \y} \[ 
	\log \L_{\blambda}(\y, \bbeta, \btau) | \blambda^{(k)} \]
	\end{equation}
	converges to a local maximum of the marginal likelihood. The difficulty herein
	is in the calculation of the expected joint log likelihood. 
	\cite{casella_empirical_2001} suggests to approximate the expectation by its Monte Carlo expectation. Although elegant and simple, this method requires a converged MCMC sample from the posterior for every iteration: a computationally intensive procedure. \cite{roy_selection_2017} introduce generalised importance sampling for the Bayesian elastic net, such that we just need a limited, pre-specified number of MCMC chains. However, this still requires several converged chains, so is not feasible in many high-dimensional omics settings. We propose to tackle this problem by approximating the expectation in (\ref{eq:ebEM}) using variational Bayes.

@@

	\section{Application to metabolomics in alzheimer}\label{sec:app}
	\subsection{Co-data separately included}
<<lines_metabolomics_alzheimer_res1_auc, fig.cap="Cross-validated AUC in alzheimer example using metabolomics class co-data, with a) the full range and b) cut-out", out.width="100%", fig.asp=2/3>>=
@

<<barplots_metabolomics_alzheimer_res1_auc, fig.cap="Estimated penalty multipliers in alzheimer example using metabolomics class co-data", out.width="50%", fig.asp=1>>=
@

@@

<<barplots_metabolomics_alzheimer_res2_auc, fig.cap="Estimated penalty multipliers in alzheimer example using quality score co-data", out.width="50%", fig.asp=1>>=
@

@@

<<barplots_metabolomics_alzheimer_res3_auc, fig.cap="Estimated penalty multipliers in alzheimer example using node degree co-data", out.width="50%", fig.asp=1>>=
@

\subsection{Co-data included simultaneously}
<<lines_metabolomics_alzheimer_res4_auc, fig.cap="Cross-validated AUC in alzheimer example using all co-data, with a) the full range and b) cut-out", out.width="100%", fig.asp=2/3>>=
@
